---
title: "Stroke-prediction-model"
author: "Faezeh Jahanshiri"
date: "2025-10-07"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    number_sections: true
    css: report.css
params:
  data_path: "C:/Users/faeze/Desktop/R-projects/healthcare-dataset-stroke-data.csv"
  id_col: "id"
  target: "stroke"
  model: "rf"          # "rf" or "logit"
  use_smote: true
  rf_trees: 800
  rf_class_weight: 10  # weight for positive class ("yes")
  train_prop: 0.8
  cv_folds: 5
  threshold: 0.25
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
set.seed(42)


if (!file.exists("report.css")) {
  writeLines(c(
    "/* --- Light styling for the HTML report --- */",
    "pre code { white-space: pre-wrap; font-size: 0.9rem; }",
    "table { width: 100%; border-collapse: collapse; }",
    "th, td { padding: 8px 10px; border-bottom: 1px solid #eee; }",
    "thead th { background: #f7f7f7; }",
    "h1, h2, h3 { margin-top: 1.4em; }",
    "code { background: #f5f5f5; padding: 2px 4px; border-radius: 4px; }"
  ), "report.css")
}

needed <- c("tidymodels","themis","janitor","vip","ranger","glmnet",
            "dplyr","ggplot2","readr","stringr")
missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing packages: ", paste(missing, collapse = ", "),
       "\nInstall once in Console, e.g.\n  install.packages(c(\"",
       paste(missing, collapse = "\", \""), "\"), lib = \"C:/Rlibs\", type = \"binary\")")
}

library(tidymodels)
library(themis)
library(janitor)
library(vip)
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)

metric_fun <- yardstick::metric_set(roc_auc, pr_auc, accuracy,
                                    yardstick::sens, yardstick::spec)

# parallel tuning
if (requireNamespace("future", quietly = TRUE)) {
  future::plan(future::multisession, workers = max(1, parallel::detectCores() - 1))
}
ctrl <- control_grid(parallel_over = "resamples", verbose = TRUE, save_pred = FALSE)



```{r load-data}
data <- read.csv(params$data_path, stringsAsFactors = FALSE)

names(data) <- names(data) |>
  tolower() |>
  gsub("[^a-z0-9]+", "_", x = _) |>
  sub("^_", "", x = _) |>
  sub("_$", "", x = _)

nrow(data); ncol(data)
head(data, 6)
str(data)
summary(data)


if ("bmi" %in% names(data)) data$bmi <- suppressWarnings(as.numeric(data$bmi))

fac_cols <- c("gender","ever_married","work_type","residence_type","smoking_status")
for (nm in intersect(fac_cols, names(data))) data[[nm]] <- factor(data[[nm]])

stopifnot(params$target %in% names(data))
s <- data[[params$target]]
if (is.factor(s)) s <- as.character(s)
if (is.character(s)) {
  s_num <- suppressWarnings(as.numeric(s))
  if (!all(s %in% c("0","1")) && any(is.na(s_num))) {
    s_num <- ifelse(tolower(s) %in% "yes", 1,
                    ifelse(tolower(s) %in% "no", 0, NA_real_))
  }
} else if (is.numeric(s)) {
  s_num <- s
} else stop("Unexpected type for target column.")

keep <- which(s_num %in% c(0,1))
stopifnot(length(keep) > 0)
data <- data[keep, , drop = FALSE]
data[[params$target]] <- factor(s_num[keep], levels = c(0,1), labels = c("no","yes"))

cat("Overall class counts:\n"); print(table(data[[params$target]]))



```{r split}
split_obj <- initial_split(data, prop = params$train_prop, strata = all_of(params$target))
train <- training(split_obj); test <- testing(split_obj)

cat("Train:", nrow(train), " Test:", nrow(test), "\n")
cat("Train classes:\n"); print(table(train[[params$target]]))
cat("Test  classes:\n"); print(table(test [[params$target]]))



```{r recipe}
rec <- recipe(reformulate(termlabels = setdiff(names(train), params$target),
                          response = params$target),
              data = train)

if (nzchar(params$id_col) && params$id_col %in% names(train)) {
  rec <- rec %>% update_role(all_of(params$id_col), new_role = "id")
}

rec <- rec %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01, other = "other") %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

if (isTRUE(params$use_smote)) rec <- rec %>% step_smote(all_outcomes())

pp <- prep(rec); p <- ncol(juice(pp)) - 1
cat("Predictors after preprocessing:", p, "\n")



```{r models, cache=TRUE}
# ---- safer, faster tuning settings ----
folds <- vfold_cv(train, v = 3, strata = all_of(params$target))

log_spec <- logistic_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>% set_mode("classification")

rf_spec_tune <- rand_forest(mtry = tune(), min_n = tune(), trees = params$rf_trees) %>%
  set_engine("ranger", importance = "impurity",
             class.weights = c("no" = 1, "yes" = as.numeric(params$rf_class_weight))) %>%
  set_mode("classification")

log_wf <- workflow() %>% add_model(log_spec) %>% add_recipe(rec)
rf_wf  <- workflow() %>% add_model(rf_spec_tune) %>% add_recipe(rec)

# compute p (predictor count) robustly
pp_try <- try(prep(rec), silent = TRUE)
p <- if (!inherits(pp_try, "try-error")) ncol(juice(pp_try)) - 1 else max(2L, ncol(train)-1)

# compact grids that avoid edge cases
log_grid <- grid_regular(penalty(range = c(-3, -0.5)), levels = 8)

rf_params <- dials::parameters(
  dials::mtry(range = c(max(2L, floor(0.1 * p)), max(2L, min(p, floor(0.6 * p))))),
  dials::min_n(range = c(2L, 15L))
)
rf_grid <- dials::grid_space_filling(rf_params, size = 8)

# reuse ctrl from setup if it exists, else make a local one
if (!exists("ctrl")) ctrl <- control_grid(parallel_over = "resamples", verbose = TRUE, save_pred = FALSE)

# ---- try tuning; if it fails, fall back automatically ----
tune_ok <- TRUE
res <- NULL
final_wf <- NULL

if (identical(params$model, "logit")) {
  res <- try(tune_grid(log_wf, resamples = folds, grid = log_grid,
                       metrics = metric_fun, control = ctrl), silent = TRUE)
} else {
  res <- try(tune_grid(rf_wf,  resamples = folds, grid = rf_grid,
                       metrics = metric_fun, control = ctrl), silent = TRUE)
}

# helper to check if tuning produced any metrics
has_metrics <- function(x) {
  inherits(x, "tune_results") &&
    tryCatch(nrow(collect_metrics(x)) > 0, error = function(e) FALSE)
}

if (!has_metrics(res)) {
  tune_ok <- FALSE
  message("Tuning failed or produced no metrics; falling back to a default Random Forest.")

  # default (no-tune) RF spec
  mtry_default <- max(2L, floor(sqrt(p)))
  rf_spec_default <- rand_forest(mtry = mtry_default, min_n = 5, trees = max(200L, as.integer(params$rf_trees))) %>%
    set_engine("ranger", importance = "impurity",
               class.weights = c("no" = 1, "yes" = as.numeric(params$rf_class_weight))) %>%
    set_mode("classification")

  final_wf <- workflow() %>% add_model(rf_spec_default) %>% add_recipe(rec)
} else {
  best <- try(select_best(res, metric = "roc_auc"), silent = TRUE)
  if (inherits(best, "try-error") || is.null(best) || nrow(best) == 0) {
    tune_ok <- FALSE
    message("No best parameters could be selected; using default Random Forest.")

    mtry_default <- max(2L, floor(sqrt(p)))
    rf_spec_default <- rand_forest(mtry = mtry_default, min_n = 5, trees = max(200L, as.integer(params$rf_trees))) %>%
      set_engine("ranger", importance = "impurity",
                 class.weights = c("no" = 1, "yes" = as.numeric(params$rf_class_weight))) %>%
      set_mode("classification")

    final_wf <- workflow() %>% add_model(rf_spec_default) %>% add_recipe(rec)
  } else {
    final_wf <- if (identical(params$model, "logit")) {
      finalize_workflow(log_wf, best)
    } else {
      finalize_workflow(rf_wf, best)
    }
  }
}

# Show top metrics if tuning worked (so you see something in the report)
if (tune_ok) {
  print(
    collect_metrics(res) %>%
      arrange(.metric, desc(mean)) %>%
      dplyr::slice_head(n = 10)
  )
} else {
  cat("Using default Random Forest (no tuning).\n")
}


```{r check-objs}
# What exists so far?
print(ls(pattern = "final_wf|final_fit|probs"))

if (!exists("final_wf")) stop("`final_wf` was not created in the [models] chunk.")

# If fit-eval didn’t run to completion, final_fit/probs might be missing.
if (!exists("final_fit")) {
  message("`final_fit` missing; fitting now...")
  final_fit <- fit(final_wf, train)
}

if (!exists("probs")) {
  message("`probs` missing; predicting now from final_fit...")
  probs <- predict(final_fit, test, type = "prob") %>%
    dplyr::bind_cols(test %>% dplyr::select(all_of(params$target)))
}

# Sanity check: we expect a .pred_yes column
stopifnot(".pred_yes" %in% names(probs))



```{r curves, fig.width=6, fig.height=4}
roc_df <- yardstick::roc_curve(probs, truth = !!sym(params$target), .pred_yes)
pr_df  <- yardstick::pr_curve(probs,  truth = !!sym(params$target), .pred_yes)

p1 <- ggplot(roc_df, aes(1 - specificity, sensitivity)) +
  geom_line() + geom_abline(lty = 2) +
  labs(x = "1 - Specificity", y = "Sensitivity")

p2 <- ggplot(pr_df, aes(recall, precision)) +
  geom_line() +
  labs(x = "Recall", y = "Precision")

p1; p2



```{r vip, fig.width=6, fig.height=5, eval = identical(params$model, "rf")}
rf_fit <- try(extract_fit_parsnip(final_fit)$fit, silent = TRUE)
if (!inherits(rf_fit, "try-error")) {
  vip::vip(rf_fit, num_features = 12)
}


```{r save-preds, eval=FALSE}
pred_out <- test %>%
  mutate(.pred_yes = probs$.pred_yes,
         .pred_no  = probs$.pred_no,
         .pred_class = factor(ifelse(.pred_yes >= params$threshold, "yes", "no"),
                              levels = c("no","yes")))
readr::write_csv(pred_out, paste0("predictions_", format(Sys.Date()), ".csv"))



sessionInfo()






